---
title: "Support Vector Machine Classifier"
author: "Clinton"
date: '2022-05-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries  
e1071 package has svm function that we can use to build a support vector machine model and the caret package will help us access functions for data partitioning and model evaluation.   

```{r}
library(e1071)
library(caret)
library(dplyr)
library(kernlab)
```

## Import the data   

```{r}
svc_data <- read.csv("./../data/student_performance.csv")
```

## Data inspection.  
Explore the data to know what data types features are of and if they are correct.  
Number of features and samples. 

```{r}
str(svc_data)
```

## Transform the data. 
* Encode the categorical variables to numeric type 

```{r}
svc_data <- svc_data %>% mutate(address = case_when(
  address == "R" ~ 0,
  address == "U" ~ 1
),
famsize = case_when(
  famsize == "LE3" ~ 0,
  famsize == "GT3" ~ 1
),
Pstatus = case_when(
  Pstatus == "T" ~ 0,
  Pstatus == "A" ~ 1
),
Mjob = case_when(
  Mjob == "teacher" ~ 0,
  Mjob == "at_home" ~ 1,
  Mjob == "services" ~ 2,
  Mjob == "other" ~ 3,
  Mjob == "health" ~ 4
),
Fjob = case_when(
  Fjob == "teacher" ~ 0,
  Fjob == "at_home" ~ 1,
  Fjob == "services" ~ 2,
  Fjob == "other" ~ 3,
  Fjob == "health" ~ 4
),
guardian = case_when(
  guardian == "mother" ~ 0,
  guardian == "father" ~ 1,
  guardian == "other" ~ 2
),
schoolsup = case_when(
  schoolsup == "yes" ~ 1,
  schoolsup == "no" ~ 0
),
famsup = case_when(
  famsup == "yes" ~ 1,
  famsup == "no" ~ 0
),
paid = case_when(
  paid == "no" ~ 0,
  paid == "yes" ~ 1
),
activities = case_when(
  activities == "no" ~ 0,
  activities == "yes" ~ 1
),
nursery = case_when(
  nursery == "no" ~ 0,
  nursery == "yes" ~ 1
),
higher = case_when(
  higher == "yes" ~ 1,
  higher == "no" ~ 0
),
internet = case_when(
  internet == "no" ~ 0,
  internet == "yes" ~ 1
),
romantic = case_when(
  romantic == "no" ~ 0,
  romantic == "yes" ~ 1
),
sex = case_when(
  sex == "F" ~ 0,
  sex == "M" ~ 1
),
G3 = case_when(
  G3 <= 10 ~ 0,
  G3 > 10 ~ 1
)
) 
```

# Drop unnecessary columns  

```{r}
svc_data <- svc_data %>% select(-c("G1","reason","school","X"))  
```


## Data partitioning 
We make use of createDataPartition function from caret package to split the data into training and testing set.  
I prefer the 80% 20% partitioning. You could do 70% 30% if you like or any proportion of your liking. 

```{r}

set.seed(123)
indexsvc = createDataPartition(svc_data$G3, p = .8, list = F)
trainsvc = svc_data[indexsvm, ]
testsvc = svc_data[-indexsvm, ]

```

## modelling    
Here we define the svm model with the default parameters. The default kernel is the radial but you can change it to linear, polynomial or sigmoid.  

```{r}
modelsvc <- e1071::svm(G3 ~ ., data = trainsvc,
                       type = 'C-classification',
                       kernel = 'linear')
summary(modelsvc)
```


## model evaluation   
```{r}
y_pred <- predict(modelsvc,testsvc)
```

```{r}
table(y_pred,testsvc$G3)
```

```{r}
correct_pred <- y_pred == testsvc$G3

prop.table(table(correct_pred))
```

